---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# commecometrics

<!-- badges: start -->
<!-- badges: end -->

commecometrics provides tools for conducting trait-based ecometric modeling with modern and fossil data. It supports workflows for both continuous and categorical environmental variables, as well as sensitivity and transferability analysis.

## Installation

You can install the development version of commecometrics from [GitHub](https://github.com/) with:

```{r}
# Install devtools if you don't have it
install.packages("devtools")

# Install commecometrics from GitHub
devtools::install_github("mariahm1995/commecometrics")
```
## Example

This is a basic example of how use commecometrics

```{r example, eval = FALSE}
library(commecometrics)

# Load example data
data("points", package = "commecometrics")
data("traits", package = "commecometrics")
data("polygons", package = "commecometrics")

# Summarize trait values at points
traitsByPoint <- summarize_traits_by_point(
  points_df = points,
  trait_df = traits,
  species_polygons = polygons,
  trait_column = "RBL",
  species_name_col = "sci_name"
)

# Run ecometric model
ecoModel <- ecometric_model(
  points_df = traitsByPoint$points,
  env_var = "BIO12",
  transform_fun = function(x) log(x + 1),
  inv_transform_fun = function(x) exp(x) - 1,
  min_species = 3
)

# Plot trait–environment surface
ecoPlot <- ecometric_space(ecoModel, env_name = "Precipitation (loge mm)")
print(ecoPlot)
```

## Learn More

See the vignette for a full walkthrough:

```{r cars}
vignette("Introduction to commecometrics")
```
